### 2022-09-06 AISTATS2023/paper.tex
started Related Work

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}

@@

@@

\section{THE U-DALE METHOD}

@@

The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.

@@

@@

Furthermore, we automate the step of splitting the axis into non-overlapping intervals. The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set. Therefore, there is an implicit trade-off behind the formation of bins. Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance). On the other hand, each bin should include points with similar local effects. Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function. We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation. Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.

>>>>

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\DY}{\mathbf{\Delta Y}}

@@

Some notation for describing the methods afterwards.

It is crucial for feature effect methods to inform about the
heterogeneous effects. Elaborate.

There are two established feture effect methods PDPlots and ALE. ALE
has some important advantages. Elaborate.

Interpretation of the heterogeneous effects behind the global effect
is available only for PDP, with three different approaches; (a) ICE
and d-ICE plots provide a visual understanding of the heterogeneous
effects. (b) grouping of ICE in homogeneous clusters, for spliting the
input space into subspace(s) with homogeneous effects (c) Feature
Interaction strength indexes, like H-statistic, provide a value
indicating how much a feature interacts with the others (not the type
of interaction).

There is no method for quantifying the heterogeneous effects, based on
ALE. Therefore, no method to exploit the advantages of ALE while, on
the same time, informing about the heterogeneous effects. We present
it in the next section.

ALE also has the peculiarity of spliting the axis into intervals,
allocating the instances of the training set in the intervals and
compute a single (constant) effect in each interval. With DALE
extension, bin spliting is decoupled from instant effect
estimation. With our extension for measuring the heterogeneous
effects, we transfrom interval spliting from a step to a clustering
problem with a meaningful objective to minimise. We provide a thorough
analysis, where we show that our objective has a consistent
meaning. It can be split in two parts; the first part is the
unavoidable uncertainty due to the natural characteristics of the
experiment, i.e.,~the data generating distribution and the black-box
function. The second part is an added uncertainty due to the
limited-samples estimation, that enforces to create groups with
constant main effect. We opt for minimizing the objective, i.e. sum of
the two uncertainties, that given that the first uncertainty is
independent of the bin spliting, therefore we want to minimize the
added uncertainty. To conclude, we transform the axis-spliting into an
unsupervised clustering problem with a principled objective. We a
computationally-grounded solution that works out-of-the-box, relaxing
the user from providing a hyperparameter without any indication which
one is the correct. This step can be used independently of whether the
user wants to explore the heterogeneous effects or not.

\section{OUR METHOD}

@@

ALE is a state-of-the-art method when it comes to feature effect
estimation. Unfortunately, there is no method for quantifying the
heterogeneous effects, based on ALE.

% The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.

@@

% \paragraph{Effect at point \(x_s\).} In the intro, we described the
% \textit{uncertainty of the global explanation} as a metric of how
% certain we are that the global explanation is valid if applied to a
% randomly-drawn individual.
% ALE plots measure the \(s\)-th feature
% effect at point \(x_s\) as the \textbf{expected} change in the output \(y\), if
% we slightly change the value of the feature of interest \(x_s\):

% \begin{equation}
%   \label{eq:ALE_mean}
%   \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
% \end{equation}

% \noindent
% We model the s-th feature effect at point \(x_s\) (change in the output \(y\) wrt a slight change in the feature of interest \(x_s\)) as the random variable $\DY ; x_s $.
% For notation convenience, we will refer to the s-th feature effect as \(\DY\), ommiting the \(;xs\) part.
% The randomness has its origins in the ignorance of the values of the rest of the features, denoted with \(\mathbf{X}_c\).
% Therefore, the s-th feature effect is defined as:

% \begin{equation}
%   \label{eq:fe_rv}
%   \DY = g(x_s) = \dfdx (x_s, \mathbf{X}_c)
% \end{equation}
% %
% As shown in Eq.~\eqref{eq:ALE_mean}, ALE is only interested in the expected value of \(\DY\).
% Instead, we are also interested in the variance of \(\DY\) for measuring the uncertainty of the local change:

% \begin{equation}
%   \label{eq:ALE_var}
%   \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
% \end{equation}

% \noindent
% The variance in Eq.~\eqref{eq:ALE_var} informs us about the heterogeneous
% effects hiding behind the explanation.

% \paragraph{Effect at interval \([z_{k-1}, z_k]\).}

% In real scenarios, we have ignorance about the generative distribution \(p(x_s, \mathbf{x}_c)\),
% residing to Monte-Carlo approximations of \(\mu(x_s)\), \(\sigma^2(x_s)\) using the samples of the training set.
% Therefore, it is impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
% point \(x_s\)\, since the  possibility to observe a sample in the interval
% \([x_s - h, x_s + h]\) is zero, in the limit where \(h \to 0\).
% Therefore, we are obliged to estimate the local effect in larger intervals (\(h>0\)).
% We refer to the expected value and the variance of the feature effect at the interval \([z_{k-1}, z_k)\), as:

% \begin{equation}
%   \label{eq:mu_bin}
%   \mu_k = \mu(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
%   \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
% \end{equation}

% \noindent

% \begin{equation}
%   \label{eq:var_bin}
%   \sigma^2_k = \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
%   \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
%       f}{\partial x_s}(x_s, \xc) - \mu_k )^2 \right] \partial z
% \end{equation}

% \noindent
% We prove that the interval-variance \(\sigma^2_k \) is:

% \begin{equation}
%  \sigma^2_k = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \sigma^2(z) + \rho^2(z) \partial z
% \end{equation}
% %
% where \(\rho(x_s) = \mu(x_s) - \mu_k\).
% The proof is at Section \ref{sec:proofs}.
% We observe that the interval-variance is the mean point-variance of the points
% inside the interval \([z_{k-1}, z_k]\), plus the mean of the residual term \(\rho\).
% The mean point-variance is the unavoidable variance, due to the uncertainty of the global explanation.
% The mean of the residual term is an extra variance (uncertainty) due to limiting the granularity of the effect at
% the bin level.

% \noindent

% \paragraph{Approximation of effect at interval \([z_{k-1}, z_k]\).}
% Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} are
% approximated by the instances of the training set that lie inside the
% \(k\)-th interval, i.e.
% \( \mathcal{S}_k = \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_{k} \} \):

% \begin{equation}
%   \label{eq:mean_estimation}
%   \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
%     \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) \right ]
% \end{equation}

% \begin{equation}
%   \label{eq:variance_estimation}
%   \hat{\sigma}_k(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
%   \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_{k-1}, z_k) \right ]^2
% \end{equation}

% \paragraph{Uncertainty of the global effect.}

% Eq.~\eqref{eq:variance_estimation} gives an approximation of the
% uncertainty of the bin effect.
% The uncertainty of the global effect is
% simply the sum of the uncertainties in the bin effects.
% The
% approximation is unbiased only if the points are uniformly distributed
% in \([z_{k-1}, z_k]\). (TODOs: Check what happens otherwise).

% \paragraph{Minimizing the uncertainty}

% Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:

% \begin{equation}
%   \label{eq:1}
%   \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
% \end{equation}
% %
% The constraints are that all bins must include more than \(\tau\)
% points, i.e., \(|\mathcal{S}_k| \geq \tau\).

% \noindent
% TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)

% \paragraph{Uncertainty of the approximation.}

% In all experiments, it also important to measure the uncertainty of
% the approximation.
% The uncertainty of the approximation can be
% quantified with two approaches:

% \begin{itemize}
% \item Splitting the dataset in many folds and (re)estimating
%   \(\hat{\mu}(z_{k-1}, z_k), \hat{\sigma}_k(z_{k-1}, z_k)\)
% \item Using the central limit theorem, we can (under assumptions) say
%   that the standard error of the approximation in
%   eq.~\eqref{eq:mean_estimation} is
%   \(\mathrm{std\_error} =
%   \frac{\hat{\sigma}_k}{\sqrt{|\mathcal{S}_k|}}\).
% \end{itemize}

@@

% Furthermore, we automate the step of splitting the axis into non-overlapping intervals. The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set. Therefore, there is an implicit trade-off behind the formation of bins. Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance). On the other hand, each bin should include points with similar local effects. Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function. We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation. Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.
